---
title: "PML project"
author: "Ganna Kaplun"
date: "4 Aug 2019"
output:
  pdf_document: default
  html_document: default
---

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now
possible to collect a large amount of data about personal activity relatively
inexpensively. These type of devices are part of the quantified self movement
- a group of enthusiasts who take measurements about themselves regularly to
improve their health, to find patterns in their behavior, or because they are
tech geeks. One thing that people regularly do is quantify how much of a
particular activity they do, but they rarely quantify how well they do it. In
this project, we use data from accelerometers on the belt,
forearm, arm, and dumbell of 6 participants.
They were asked to perform barbell lifts correctly and incorrectly in 5
different ways. Given data from accelerometers, the goal is to predict the class
of action which is one of the following.

- exactly according to the specification (A)
- throwing elbows to the front (B)
- lifting the dumbbell only halfway (C)
- lowering the dumbbell only halfway (D)
- throwing the hips to the front (E).


More information is available from the website here: 
http://groupware.les.inf.puc-rio.br/har
(see the section on the Weight Lifting Exercise Dataset). 

# Data

Here is the training data for this project: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

Here is the test data: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

```{r cache=TRUE}
library(RCurl)
train_uri <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_uri <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
train_data <- read.csv(text=getURL(train_uri), na.strings=c("", "NA"))
test_data <- read.csv(text=getURL(test_uri), na.strings=c("", "NA"))
```

The index is the first column of the data which is of no use. Let's remove it from training data.
```{r}
train_data$X <- NULL
```

Correspondingly, it is extremely unlikely that the user and time information have any effect on if barbell lifts are performed correctly.

```{r}
cols_remove <- c("user_name", "raw_timestamp_part_1",
                    "raw_timestamp_part_2", "cvtd_timestamp")
for (col in cols_remove) {
    train_data[, col] <- NULL
}
```

Lots of columns in the dataset include numerous missing values. Let's extract
features from both data sets with too many missing values, where imputing is not possible.

```{r}
NAs <- apply(train_data,2,function(x) {sum(is.na(x))})
train_data <- train_data[,which(NAs == 0)]
```

Let's remove features without many missing values but only one unique value (i.e. zero variance predictors) or with few unique values compared to the number
of samples and those where the frequency ratio of the most common value to the
frequency of the second most common value is significant.

```{r message=FALSE}
library(caret)
nsv <- nearZeroVar(train_data)
train_data <- train_data[-nsv]
test_data <- test_data[-nsv]
```

The final collection of predictors for classification is the following.

```{r}
names(train_data)
```

# The model

The first one is a random forest classifier to predict the action class. To evaluate the model accuracy, let's perform 10-fold crossvalidation with 80:20 split, i.e. on each fold 80% of the data is applied for training the random forest and the rest 20% is for testing.

```{r cache=TRUE}
library(randomForest)
set.seed(1)
obs <- c()
preds <- c()
for(i in 1:10) {
    intrain = sample(1:dim(train_data)[1], size=dim(train_data)[1] * 0.8, replace=F)
    train_cross = train_data[intrain,]
    test_cross = train_data[-intrain,]
    rf <- randomForest(classe ~ ., data=train_cross)
    obs <- c(obs, test_cross$classe)
    preds <- c(preds, predict(rf, test_cross))
}
```

The confusion matrix for predictions on crossvalidation folds is below.

```{r}
conf_mat <- confusionMatrix(table(preds, obs))
conf_mat$table
```

The proposed model seems to do well enough. The accuracy is 
`r conf_mat$overall[[1]] * 100`% and it misclassifies only few instances. 
Last, let's train the random forest
with all the dataset so that the classifier could be used to predict the class of an action, given the set of activity measurements.

```{r cache=TRUE}
model <- randomForest(classe ~ ., data=train_data)
```

